---
title: "In Class Exercise October 20"
author: "Lucy Cousins"
date: "10/20/2020"
output: 
  html_document: 
    toc: TRUE
    toc_float: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##Making a new chunk
Windows - control+alt+i
Macs - command+option+i

I can give the code chunk a name by adding words in the curly braces


## Reading a csv file

I'm going to read some data from a csv (comma-separated-values) file.

I can use read.csv() along with the path to the file to read the data into a data.frame.

##Here's an example:

```{r read_birds_data}
require(here)
dat_birds = read.csv(here("data", "hab.sta.csv"))
dat_birds <- read.csv(here("data", "hab.sta.csv"))
```

## Previewing the contents of a data.frame

I can use the head() function to print the first few lines of a data.frame.
Here's an example with the birds data

```{r head_function_demo}
head(dat_birds)
```
##Histogram
I will use hist() to create a historgram 
```{r simple_plot}
hist(dat_birds$aspect, col=5)
```

##Fancy Histogram
```{r custom histogram 1}
hist(
  dat_birds$aspect, 
  col=5, 
  main='Aspect Histogram', 
  xlab='Aspect',
  breaks=18
  )

```

